Emergent dual scaling of riverine biodiversity

### 2021-07-02 manuscript_pnas.Rmd
Update manuscript_pnas.Rmd

The past two decades have witnessed efforts to illuminate the role of spatial ecosystem structure in shaping biodiversity patterns [@rinaldoRiverNetworksEcological2018; @grantLivingBranchesPopulation2007]. Yet, current schemes have overlooked one of the fundamental phenomena in spatially complex ecosystems - the dual scaling law. Our unified framework unveiled the surprising consistency of dual scaling across ecological contexts or geographic regions, suggesting the universality of our findings. This discovery, therefore, should bring a crucial perspective to macroecological science.

In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

While prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in stochastic/deterministic community assembly [@chaseStochasticCommunityAssembly2010]. Recognizing the dual scaling of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through the Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through the Water Quality Benefits grant (EPA-G2015-STAR-A1), the University of North Carolina at Greensboro through start-up funds, and the Japan Society for the Promotion of Science through the KAKENHI grant (20H03010).

>>>>

Our unified framework unveiled the surprising consistency of dual scaling across ecological contexts or geographic regions, suggesting the universality of our findings. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020]. Recognizing the dual scaling of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it.

Ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. However, while the foundational concepts in ecology emphasize the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. The ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. In this context, our study serves as a conceptual pillar for future research. For example, the proposed framework may be extended to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002] and shifts in stochastic/deterministic community assembly [@chaseStochasticCommunityAssembly2010].The discovery of the present study should therefore bring a paradigm shift in macroecological science.

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Illinois Environmental Protection Agency, Minnesota Pollution Control Agency, Iowa Department of Natural Resources, Wisconsin Department of Natural Resources, Michio Fukushima and Hirokazu Urabe for providing data for our empirical analysis. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through the Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through the Water Quality Benefits grant (EPA-G2015-STAR-A1), the University of North Carolina at Greensboro through start-up funds, and the Japan Society for the Promotion of Science through the KAKENHI grant (20H03010).
